Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-01-09 manuscript/manuscript.tex
Conversion to Quarto file

---
title: "Flattering Advice: Avoiding Disappointment in Advice-Giving"
shorttitle: "Flattering Advice"
date: yes
author: 
  - name: "Amanda Chen"
    corresponding: true
    orcid: 0000-0002-4428-627X
    email: zchengj@connect.ust.hk
    affiliations:
      - id: id1
        name: "The Hong Kong University of Science and Technology"
        department: "Department of Management"
        city: "Hong Kong"
  - name: "David Hagmann"
    orcid: 0000-0002-2080-997X
    affiliations:
    - id: id1
author-note:
  disclosures:
    gratitude: The authors are grateful to XXX for their helpful comments.
abstract: Good advice improves decision quality but often requires delivering unpleasant truths that may disappoint advisees. Across three pre-registered and incentivized experiments involving real adviser-advisee interactions (N = 3,900), we show that advisers prioritize avoiding disappointment at the expense of accuracy and their own earnings. In Study 1, advisers financially rewarded for accuracy still tailor recommendations to aspirational goals expressed by advisees, resulting in worse advice. When incentivized to be liked, advisers provide even more flattering advice, and advisees reward this by rating these advisers as more likable, despite the advice being less honest and less accurate (Study 2). The desire to avoid disappointment may lead to inequities if there are differences in expectations across social groups. In Study 3, we examine a setting in which men expect to perform better than women. We show that advisers take into account these expectations, leading to systematically different advice for men and women even when their gender is unknown to advisers. Advisers' efforts to avoid disappointment may thus contribute to systematic gender disparities in advice, with implications for downstream decision-making.

keywords: [Gender difference, Advice, Interpresonal relationship]
bibliography: bibliography.bib
floatsintext: true
mask: false
link-citations: false
format:
  apaquarto-html: default
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    prefer-html: true
  apaquarto-docx:
    default: true
    prefer-html: true
---

```{r}
#| label: setup
#| include: false

library(magrittr)
library(ltm)
library(ggplot2)
library(gridExtra)
library(dplyr)

knitr::opts_chunk$set(fig.width = 8, fig.height = 4, cache = F, echo = F,
                      fig.align = 'center',warning = FALSE, message = FALSE)

study1_time1 <- readRDS(here::here("Data/Study1_time1.Rds"))
study1_ability50 <- read.csv(here::here("Data/Cleaned Data/S1_Ability50_clean.csv"))
study1_time2_giver <- readRDS(here::here("Data/Study1_time2_giver.Rds"))
study1_time2_advice <- readRDS(here::here("Data/Study1_time2_advice.Rds"))
study1_time3 <- readRDS(here::here("Data/Study1_time3.Rds"))

```

# Introduction

Career trajectories for men and women often differ substantially, even among those with similar qualifications. One source of such differences may be the advice that people receive. Prior work has documented that men receive more aspirational advice and women more risk-averse advice, which may explain why women are less likely to apply for and ultimately obtain more rewarding positions (Kanze et al., 2018). Research on gender discrimination has focused on the role of unconscious bias (Greenwald et al., 2009; Greenwald & Banaji, 1995), but interventions that try to reduce this bias have been largely unsuccessful (Chang et al., 2019; Paluck & Green, 2009). Recent work has proposed cognitive, rather than unconscious, biases that can lead to gender discrimination through the formation of false beliefs (Hagmann et al., 2024). Here, we propose a novel source of differences in advice by drawing on recent work on belief-based utility, expectation-based reference points (Köszegi, 2006; Kőszegi & Rabin, 2009), and the hedonic consequences of information (Golman et al., 2017).

Advice figures prominently in workplace and personal decision-making, shaping choices about promotions, career paths, and everyday dilemmas [@Gino2007; @Harvey1997; @Soll2009]. Its fundamental premise is to improve outcomes by enabling decision-makers to learn from others' expertise. However, while honest advice can lead to better decisions, it can also lead to disappointment when it suggests a desired outcome is unlikely to materialize. Advisers may be motivated to offer flattering recommendations that align with advisees' hopes, sparing them the disappointment of a more sobering forecast.

@@

People often go to great lengths to avoid unpleasant information, deriving belief-based utility from maintaining favorable views of themselves [@Golman2017; @Ho2018; @Loewenstein2018]. Moreover, they often prefer advisers who withhold bad news from them [@Shalvi2019] and punish those who communicate bad news [@John2019]. Advisers, anticipating these preferences and consequences face a dilemma: Should they provide the accurate but potentially painful truth, or shade their recommendations to preserve the advisee's mood and the adviser's own standing? Advisers who fear they will be blamed or disliked for undermining someone's confidence may reasonably avoid candid feedback, particularly if they don't incur a cost when the advisee makes a mistake.

In our experiments, however, advisees have no opportunity to punish the adviser, and advisers' financial incentives are linked to the decisions the advisees make. We propose that advisers nonetheless have reason to provide flattering advise because they recognize the psychological toll of disappointing news. As a result, they may similarly incur a hedonic cost for delivering unpleasant information that they anticipate may distress the advisee. Thus, even in the anonymized context of an online experiment, we hypothesize (and find) that advisers are relucant to disappoint advisees and advice is thus biased upward, away from accuracy and toward confirming advisees' priors. We show that this reduces the quality of advice—but that advisees nonetheless view flattering advisers more favorably.

We present the results of three preregistered experiments in which participants are paired anonymously as advisers and advisees. Advisers are incentivized to offer accurate advice, with bonus earnings depending on the outcome of an advisee's decision. We study a setting in which advisees can communicate their expectations to an adviser, and experimentally manipulate whether advisers observe this expectation. Our design simulates organizational contexts where mentors and managers often have some sense of an employee's confidence or aspirations and thus could adjust their feedback to avoid disappointing them. Interpersonal considerations would likely be stronger in situations in which the adviser and advisee have an existing relationship and when the advice is delivered face-to-face. Moreover, advisers usually do not suffer any direct costs when their flattering advice leads to a poor outcome for the recipient, particularly when it is not clear what would have happened under a counterfactual.\^\[In some cases, however, repeated interaction might also increase incentives for honest feedback. Someone who is known to persistently give overoptimistic advice may be viewed as less trustworthy in the long run.\]

@@

In Study 3, we examine a setting in which participants report their expected performance on a mathematics quiz. Advisers observe the test-taker's true score and, in one treatment, the score they guessed they would receive. They then recommend whether the advisee should compete against a group of high performers or low performers. Importantly, this "competition" is based on the past performance observed by the adviser and thus the advisee's expectation does not provide instrumental information. However, we find that (1) men expect to perform better than women, given identical performance, and (2) advisers take these expectations into account. As a result, when expectations (but not gender) are known to advisers, the advice given to men and women differs, and men are more often advised to compete against the group of high performers. Notably, this leads to worse advice for men than for women.

# Open Science Statement

We report all manipulations, measures, and data exclusion in our experiments. The preregistration reports, screenshots of all experimental materials, and the analysis code to replicate all statistical analyses and figures are available on the Open Science Framework (<https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d>).

# Study 1

We first examine whether advisers take into account the expectations of advisees when providing a recommendation. A group of participants ("Advisees") complete a quiz consisting of ten questions that draw on ego-relevant domains. They then express a non-binding preference to compete against a group of high performers or low performers on the same task, based on their past performance. We nudge participants towards picking either the low or the high performer group by selecting one of the two by default. A second group of participants ("Advisers") observe Advisees' score on the quiz and, in one treatment, also which group they had selected. We predicted that Advisees who were nudged toward the high performer group would be more likely to receive advice to compete against this high group when Advisers observed their non-binding choice.

## Methods

In a preliminary stage, we recruited `r scales::comma(dim(study1_ability50)[1])` participants from Prolific and gave them five minutes to complete a quiz consisting of ten items. The quiz included word puzzles, identifying emotions from photos, and selecting the best responses for hypothetical scenarios. These questions were adapted from surveys that measure problem-solving, emotional intelligence, and communication skills---which we pre-tested as being important in modern society and hence where participants might have a stake in doing well. Participants received a bonus of five cents for each correctly answered question. We then ranked them based on their score and label the 20 participants with the lowest (highest) scores as the Low (High) Performer Group.

Next, we recruited `r scales::comma(dim(study1_time1)[1])` participants for the role of Advisees. They completed the same 10-item quiz, also earning five cents for each correct answer. We then asked them to express a non-binding preference for whether they would like to compete against the High or Low Performer Group. We informed them that they would be invited back at a later date when they could make a binding decision and they could earn a bonus based on whether their score on this quiz was equal to or higher than a randomly selected member from their comparison group. If they picked the High Performer Group and had an equal or higher score than a randomly selected member of that group, they would earn a bonus of 50 cents. If they picked the Low Performer Group and outperformed a randomly selected member of that group, they could earn a bonus of 20 cents. If their score was lower in this comparison, they would not earn an additional bonus. We randomized which of the two groups (High or Low Performers) was selected by default, and Advisees were free to select the other group. They also made a guess as to their score, and the survey concluded with basic demographic questions.

We then recruited `r scales::comma(dim(study1_time2_giver)[1])` participants for the role of Advisers, and the focal part of our experiment. They were informed of the ability quiz that the Advisees had completed, how the Low Performer Group and High Performer Groups were constructed, as well as the choice and incentives for the Advisees. We then asked Advisers to give advice to ten participants on which group they should compete against. Advisers were randomly assigned to one of two treatments. In the "Performance" treatment, they observed only the score of the other participant. In the "Performance + Expectation" treatment, they additionally observed the advisee's non-binding choice of which group to compete against. However, since the outcome was based solely on the past quiz score and advisers, but not advisees, know the score, this initial choice was not informative for the purpose of recommending competing against one group or the other. Advisers could earn the same bonus as one of the participants they had given advice to and who returned to make a decision. The survey concluded with basic demographic questions.

Finally, we invited Advisees back for the follow-up survey. Following our preregistration, we kept the survey open for 7 days. In total, `r scales::comma(dim(study1_time3)[1])` Advisees returned. The short survey reminded them of the task they completed in the previous survey, informed them that other participants from Prolific had observed their real score and given them advice on which group to compete against, and finally reminded them of how many questions they guessed they had answered correctly. They were not informed of their true score or the score of the groups they could compete against. Participants then observed the advice from a randomly selected adviser and made their decision.

## Results

```{r}
contingency_table <- table(study1_time1$Guess, study1_time1$default)

chi_test <- chisq.test(contingency_table)

```

We begin by examining the performance of the Stage 1 participants. Because our default treatment takes place after participants completed the ability quiz, we do not expect a difference in performance across the Low and High Default treatments. Indeed, the two groups scored no different from one another (`r round(mean(study1_time1[study1_time1$default == "Low",]$Score),2)` and `r round(mean(study1_time1[study1_time1$default == "High",]$Score),2)` for the Low Default and High Default treatments, respectively, `r papaja::apa_print(t.test(Score ~ default, data = study1_time1, var.equal = T))$statistic`). The default manipulation affected their initial choices. In the "Low Default" treatment, only `r round(mean(study1_time1[study1_time1$default == "Low",]$Guess),2)*100`% of participants initially preferred competing against the High Performer Group, compared to `r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100`% of participants who saw the High Performer Group highlighted ($\chi^2$(1, n = 201) = 36.62, p \< .001).

```{r}
#| label: tbl-study1regs
#| tbl-cap: "Advice to compete against the High Performer Group in Study 1. Displaying the non-binding choice of those for whom the Low Performer Group was selected by default makes it more likely that they are advised to compete against the High Performer Group (Column 1). Column 2 restricts the analysis to advisers who observed advisees' initial non-binding choice and controls for advisees' score on the quiz. Standard errors are clustered at the adviser level."

study1_advice <- study1_time2_advice %>%
  dplyr::mutate(Treatment = factor(Treatment, levels = c('Low', 'High'))) %>%
  dplyr::mutate(ExpShow = ifelse(Condition == 'E', 1, 0))

modelsummary::modelsummary(
  list(`(1)` = lm(Advice ~ ExpShow * Treatment, data = study1_advice), 
       `(2)` = lm(Advice ~ Treatment + Score, data = study1_advice %>% dplyr::filter(ExpShow == 1))),
  vcov = ~GiverID,
  coef_map = c(
    "TreatmentHigh" = "High Default",
    "ExpShow" = "Expectation Shown",
    "ExpShow:TreatmentHigh" = "Expectation x High Default",
    "Score" = "Performance",
    "(Intercept)" = "Constant"
  ),
  stars = TRUE,
  gof_map = list(list('raw' = 'nobs', 'clean' = 'N', 'fmt' = 0))
)

```

We now examine whether advisers took the advisees' expectations into account. Recall that advisees who were defaulted to competing against the High Performer Group were indeed more likely to choose this group, which, from the advisers' perspective, could signal high confidence in their abilities. Our theory predicts that showing expectations should lead to more flattering advice--- specifically, a recommendation to compete against the High Performer Group--- for advisees with a high default choice compared to those with a low default choice. Column 1 of @tbl-study1regs reports a linear probability model with advice to compete against the high group as the outcome measure, incorporating the default choice advisees saw, whether expectations were shown to the adviser, and the interaction of the two. Because each adviser made ten recommendations, we cluster standard errors at the adviser level. As predicted, we find a marginally significant interaction effect: advisees with a default choice to compete against the High Performer Group (nudged to have high expectations, though not necessarily holding them) were more likely to receive advice to compete against stronger competitors when their expectations were shown compared to when they were hidden. We reported a linear probability model on advising to compete against the High Performer group for advisers in the "Expectations" treatment, controlling for the true performance of the advisee in Column 2 of @tbl-study1regs. We observe a significant main effect of the default treatment, where a high default increased the likelihood of receiving flattering advice. It is worth noting that, since only `r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100` % of advisees in the high default group actually stuck with their default choice, this represents a conservative test of the effect of high expectations on the likelihood of receiving flattering advice.

```{r}
#| label: fig-study1Advice
#| fig-cap: "Advisees who were nudged to express high expectations were more likely to receive advice to compete against the High Performer Group."

>>>>

\documentclass[
  man,
  floatsintext,
  longtable,
  nolmodern,
  notxfonts,
  notimes,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{apa7}

\usepackage{amsmath}
\usepackage{amssymb}

\RequirePackage{longtable}
\RequirePackage{threeparttablex}

\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
	{0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
	{-.5em}%
	{\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{0.5em}%
	{0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
	{-\z@\relax}%
	{\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

\usepackage{longtable, booktabs, multirow, multicol, colortbl, hhline, caption, array, float, xpatch}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

\usepackage{tcolorbox}
\tcbuselibrary{listings,theorems, breakable, skins}
\usepackage{fontawesome5}

\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{ACACAC}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582EC}
\definecolor{quarto-callout-important-color-frame}{HTML}{D9534F}
\definecolor{quarto-callout-warning-color-frame}{HTML}{F0AD4E}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02B875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{FD7E14}

%\newlength\Oldarrayrulewidth
%\newlength\Oldtabcolsep

\usepackage{hyperref}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{newtx}

\defaultfontfeatures{Scale=MatchLowercase}
\defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}

\title{Flattering Advice: Avoiding Disappointment in Advice-Giving}

\shorttitle{Flattering Advice}

\usepackage{etoolbox}

\authorsnames{Amanda Chen,David Hagmann}

\authorsaffiliations{
{Department of Management, The Hong Kong University of Science and
Technology},{}}

\leftheader{Chen and Hagmann}

\date{Invalid Date}

\abstract{Good advice improves decision quality but often requires
delivering unpleasant truths that may disappoint advisees. Across three
pre-registered and incentivized experiments involving real
adviser-advisee interactions (N = 3,900), we show that advisers
prioritize avoiding disappointment at the expense of accuracy and their
own earnings. In Study 1, advisers financially rewarded for accuracy
still tailor recommendations to aspirational goals expressed by
advisees, resulting in worse advice. When incentivized to be liked,
advisers provide even more flattering advice, and advisees reward this
by rating these advisers as more likable, despite the advice being less
honest and less accurate (Study 2). The desire to avoid disappointment
may lead to inequities if there are differences in expectations across
social groups. In Study 3, we examine a setting in which men expect to
perform better than women. We show that advisers take into account these
expectations, leading to systematically different advice for men and
women even when their gender is unknown to advisers. Advisers' efforts
to avoid disappointment may thus contribute to systematic gender
disparities in advice, with implications for downstream
decision-making.}

\keywords{Gender difference, Advice, Interpresonal relationship}

\authornote{\par{\addORCIDlink{Amanda
Chen}{0000-0002-4428-627X}}\par{\addORCIDlink{David
Hagmann}{0000-0002-2080-997X}}

\par{     The authors are grateful to XXX for their helpful comments.  }
\par{Correspondence concerning this article should be addressed
to Amanda Chen, Email: zchengj@connect.ust.hk}
}

\makeatletter
\let\endoldlt\endlongtable
\def\endlongtable{
\hline
\endoldlt
}
\makeatother

\urlstyle{same}

\usepackage{float}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{rotating}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

% From https://tex.stackexchange.com/a/645996/211326
%%% apa7 doesn't want to add appendix section titles in the toc
%%% let's make it do it
\makeatletter
\xpatchcmd{\appendix}
  {\par}
  {\addcontentsline{toc}{section}{\@currentlabelname}\par}
  {}{}
\makeatother

%% Disable longtable counter
%% https://tex.stackexchange.com/a/248395/211326

\usepackage{etoolbox}

\makeatletter
\patchcmd{\LT@caption}
  {\bgroup}
  {\bgroup\global\LTpatch@captiontrue}
  {}{}
\patchcmd{\longtable}
  {\par}
  {\par\global\LTpatch@captionfalse}
  {}{}
\apptocmd{\endlongtable}
  {\ifLTpatch@caption\else\addtocounter{table}{-1}\fi}
  {}{}
\newif\ifLTpatch@caption
\makeatother

\begin{document}

\maketitle

\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\setlength\LTleft{0pt}

\section{Introduction}\label{introduction}

Career trajectories for men and women often differ substantially, even
among those with similar qualifications. One source of such differences
may be the advice that people receive. Prior work has documented that
men receive more aspirational advice and women more risk-averse advice,
which may explain why women are less likely to apply for and ultimately
obtain more rewarding positions (Kanze et al., 2018). Research on gender
discrimination has focused on the role of unconscious bias (Greenwald et
al., 2009; Greenwald \& Banaji, 1995), but interventions that try to
reduce this bias have been largely unsuccessful (Chang et al., 2019;
Paluck \& Green, 2009). Recent work has proposed cognitive, rather than
unconscious, biases that can lead to gender discrimination through the
formation of false beliefs (Hagmann et al., 2024). Here, we propose a
novel source of differences in advice by drawing on recent work on
belief-based utility, expectation-based reference points (Köszegi, 2006;
Kőszegi \& Rabin, 2009), and the hedonic consequences of information
(Golman et al., 2017).

Advice figures prominently in workplace and personal decision-making,
shaping choices about promotions, career paths, and everyday dilemmas
(Gino \& Moore, 2007; Harvey \& Fischer, 1997; Soll \& Larrick, 2009).
Its fundamental premise is to improve outcomes by enabling
decision-makers to learn from others' expertise. However, while honest
advice can lead to better decisions, it can also lead to disappointment
when it suggests a desired outcome is unlikely to materialize. Advisers
may be motivated to offer flattering recommendations that align with
advisees' hopes, sparing them the disappointment of a more sobering
forecast.

@@

People often go to great lengths to avoid unpleasant information,
deriving belief-based utility from maintaining favorable views of
themselves (Golman et al., 2017; Loewenstein \& Molnar, 2018;
\textbf{Ho2018?}). Moreover, they often prefer advisers who withhold bad
news from them (Shalvi et al., 2019) and punish those who communicate
bad news (John et al., 2019). Advisers, anticipating these preferences
and consequences face a dilemma: Should they provide the accurate but
potentially painful truth, or shade their recommendations to preserve
the advisee's mood and the adviser's own standing? Advisers who fear
they will be blamed or disliked for undermining someone's confidence may
reasonably avoid candid feedback, particularly if they don't incur a
cost when the advisee makes a mistake.

In our experiments, however, advisees have no opportunity to punish the
adviser, and advisers' financial incentives are linked to the decisions
the advisees make. We propose that advisers nonetheless have reason to
provide flattering advise because they recognize the psychological toll
of disappointing news. As a result, they may similarly incur a hedonic
cost for delivering unpleasant information that they anticipate may
distress the advisee. Thus, even in the anonymized context of an online
experiment, we hypothesize (and find) that advisers are relucant to
disappoint advisees and advice is thus biased upward, away from accuracy
and toward confirming advisees' priors. We show that this reduces the
quality of advice---but that advisees nonetheless view flattering
advisers more favorably.

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their feedback to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.\^{}{[}In some cases, however,
repeated interaction might also increase incentives for honest feedback.
Someone who is known to persistently give overoptimistic advice may be
viewed as less trustworthy in the long run.{]}

@@

In Study 3, we examine a setting in which participants report their
expected performance on a mathematics quiz. Advisers observe the
test-taker's true score and, in one treatment, the score they guessed
they would receive. They then recommend whether the advisee should
compete against a group of high performers or low performers.
Importantly, this ``competition'' is based on the past performance
observed by the adviser and thus the advisee's expectation does not
provide instrumental information. However, we find that (1) men expect
to perform better than women, given identical performance, and (2)
advisers take these expectations into account. As a result, when
expectations (but not gender) are known to advisers, the advice given to
men and women differs, and men are more often advised to compete against
the group of high performers. Notably, this leads to worse advice for
men than for women.

\section{Open Science Statement}\label{open-science-statement}

We report all manipulations, measures, and data exclusion in our
experiments. The preregistration reports, screenshots of all
experimental materials, and the analysis code to replicate all
statistical analyses and figures are available on the Open Science
Framework
(\url{https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d}).

\section{Study 1}\label{study-1}

We first examine whether advisers take into account the expectations of
advisees when providing a recommendation. A group of participants
(``Advisees'') complete a quiz consisting of ten questions that draw on
ego-relevant domains. They then express a non-binding preference to
compete against a group of high performers or low performers on the same
task, based on their past performance. We nudge participants towards
picking either the low or the high performer group by selecting one of
the two by default. A second group of participants (``Advisers'')
observe Advisees' score on the quiz and, in one treatment, also which
group they had selected. We predicted that Advisees who were nudged
toward the high performer group would be more likely to receive advice
to compete against this high group when Advisers observed their
non-binding choice.

\subsection{Methods}\label{methods}

In a preliminary stage, we recruited 50 participants from Prolific and
gave them five minutes to complete a quiz consisting of ten items. The
quiz included word puzzles, identifying emotions from photos, and
selecting the best responses for hypothetical scenarios. These questions
were adapted from surveys that measure problem-solving, emotional
intelligence, and communication skills---which we pre-tested as being
important in modern society and hence where participants might have a
stake in doing well. Participants received a bonus of five cents for
each correctly answered question. We then ranked them based on their
score and label the 20 participants with the lowest (highest) scores as
the Low (High) Performer Group.

Next, we recruited 201 participants for the role of Advisees. They
completed the same 10-item quiz, also earning five cents for each
correct answer. We then asked them to express a non-binding preference
for whether they would like to compete against the High or Low Performer
Group. We informed them that they would be invited back at a later date
when they could make a binding decision and they could earn a bonus
based on whether their score on this quiz was equal to or higher than a
randomly selected member from their comparison group. If they picked the
High Performer Group and had an equal or higher score than a randomly
selected member of that group, they would earn a bonus of 50 cents. If
they picked the Low Performer Group and outperformed a randomly selected
member of that group, they could earn a bonus of 20 cents. If their
score was lower in this comparison, they would not earn an additional
bonus. We randomized which of the two groups (High or Low Performers)
was selected by default, and Advisees were free to select the other
group. They also made a guess as to their score, and the survey
concluded with basic demographic questions.

We then recruited 1,000 participants for the role of Advisers, and the
focal part of our experiment. They were informed of the ability quiz
that the Advisees had completed, how the Low Performer Group and High
Performer Groups were constructed, as well as the choice and incentives
for the Advisees. We then asked Advisers to give advice to ten
participants on which group they should compete against. Advisers were
randomly assigned to one of two treatments. In the ``Performance''
treatment, they observed only the score of the other participant. In the
``Performance + Expectation'' treatment, they additionally observed the
advisee's non-binding choice of which group to compete against. However,
since the outcome was based solely on the past quiz score and advisers,
but not advisees, know the score, this initial choice was not
informative for the purpose of recommending competing against one group
or the other. Advisers could earn the same bonus as one of the
participants they had given advice to and who returned to make a
decision. The survey concluded with basic demographic questions.

Finally, we invited Advisees back for the follow-up survey. Following
our preregistration, we kept the survey open for 7 days. In total, 176
Advisees returned. The short survey reminded them of the task they
completed in the previous survey, informed them that other participants
from Prolific had observed their real score and given them advice on
which group to compete against, and finally reminded them of how many
questions they guessed they had answered correctly. They were not
informed of their true score or the score of the groups they could
compete against. Participants then observed the advice from a randomly
selected adviser and made their decision.

\subsection{Results}\label{results}

We begin by examining the performance of the Stage 1 participants.
Because our default treatment takes place after participants completed
the ability quiz, we do not expect a difference in performance across
the Low and High Default treatments. Indeed, the two groups scored no
different from one another (5.1 and 5.18 for the Low Default and High
Default treatments, respectively, \(t(199) = 0.31\), \(p = .756\)). The
default manipulation affected their initial choices. In the ``Low
Default'' treatment, only 22\% of participants initially preferred
competing against the High Performer Group, compared to 65\% of
participants who saw the High Performer Group highlighted (\(\chi^2\)(1,
n = 201) = 36.62, p \textless{} .001).

\begin{table}

{\caption{{Advice to compete against the High Performer Group in Study
1. Displaying the non-binding choice of those for whom the Low Performer
Group was selected by default makes it more likely that they are advised
to compete against the High Performer Group (Column 1). Column 2
restricts the analysis to advisers who observed advisees' initial
non-binding choice and controls for advisees' score on the quiz.
Standard errors are clustered at the adviser
level.}{\label{tbl-study1regs}}}
\vspace{-20pt}}

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]},
column{2,3}={}{halign=c,},
column{1}={}{halign=l,},
hline{12}={1,2,3}{solid, black, 0.05em},
}                     %% tabularray inner close
\toprule
& (1) & (2) \\ \midrule %% TinyTableHeader
High Default               & \num{0.011}    & \num{0.025}*    \\
& (\num{0.013})  & (\num{0.011})   \\
Expectation Shown          & \num{-0.041}*  &                  \\
& (\num{0.016})  &                  \\
Expectation x High Default & \num{0.037}+   &                  \\
& (\num{0.019})  &                  \\
Performance                &                 & \num{0.168}***  \\
&                 & (\num{0.004})   \\
Constant                   & \num{0.389}*** & \num{-0.503}*** \\
& (\num{0.012})  & (\num{0.020})   \\
N                          & \num{10000}    & \num{5000}      \\
\bottomrule
\end{talltblr}

\end{table}

We now examine whether advisers took the advisees' expectations into
account. Recall that advisees who were defaulted to competing against
the High Performer Group were indeed more likely to choose this group,
which, from the advisers' perspective, could signal high confidence in
their abilities. Our theory predicts that showing expectations should
lead to more flattering advice--- specifically, a recommendation to
compete against the High Performer Group--- for advisees with a high
default choice compared to those with a low default choice. Column 1 of
Table~\ref{tbl-study1regs} reports a linear probability model with
advice to compete against the high group as the outcome measure,
incorporating the default choice advisees saw, whether expectations were
shown to the adviser, and the interaction of the two. Because each
adviser made ten recommendations, we cluster standard errors at the
adviser level. As predicted, we find a marginally significant
interaction effect: advisees with a default choice to compete against
the High Performer Group (nudged to have high expectations, though not
necessarily holding them) were more likely to receive advice to compete
against stronger competitors when their expectations were shown compared
to when they were hidden. We reported a linear probability model on
advising to compete against the High Performer group for advisers in the
``Expectations'' treatment, controlling for the true performance of the
advisee in Column 2 of Table~\ref{tbl-study1regs}. We observe a
significant main effect of the default treatment, where a high default
increased the likelihood of receiving flattering advice. It is worth
noting that, since only 65 \% of advisees in the high default group
actually stuck with their default choice, this represents a conservative
test of the effect of high expectations on the likelihood of receiving
flattering advice.

\begin{figure}
